---
title: "R Notebook"
output: html_notebook
---

```{r}
#install.packages("lattice")
library(Metrics)
# library(ggplot2)
library(gplots)
library(lattice)
library(countcolors)
```



Utility functions, please run before analysis

```{r}
# Get lower triangle of the correlation matrix
get_lower_tri<-function(cormat){
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}
# Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
  cormat[lower.tri(cormat)]<- NA
  return(cormat)
}
  
# calculate the difference between a pattern and its Xth descendant
# then average for the whole lineage
# ie sum of the matrix rows div by count(non null entries)
average_difference<-function(df){
  # we want df[1,1:end] + df[2, 2:end]
  n = length(df[,1])
  d = df[1, ]
  # sum after moving the row back by one column
  for (i in c(2:n)){
    sub_ = df[i,i:n]
    d = d + c(sub_, rep(0, n-length(sub_)))
  }
  
  # divide by number of non null entries
  d = d/c(n:1)
 
  return(d)
}
```


Read data from file

```{r}

# download data from https://figshare.com/projects/Self-replicating_Neural_Cellular_Automata/167582, 
# or run experiments from https://github.com/LanaSina/NCA_self_replication/tree/main
# replace with correct folder
folder_name = "../alife_2023/fish/csv/1500/"

# sorry, no config file
max_step= 19200
egg_filename = paste(folder_name, "egg_", max_step,".csv", sep = "")
fish_filename = paste(folder_name, "fish_", max_step,".csv", sep = "")

df_e = read.csv(egg_filename,stringsAsFactors=F, header = FALSE)
df_f = read.csv(fish_filename,stringsAsFactors=F, header = FALSE)

```


Calculate DNA drift

```{r}
# for each egg calculate the difference with others
n = length(df_e[,1])
e_differences = matrix(ncol=n, nrow=n)

# row, col
# advances col by col so generation diff is 1st row
for ( i in c(1:n)){
  current = as.numeric(df_e[i,])
  for ( j in c(i:n)){
    egg = as.numeric(df_e[j,])
    diff = mse(current, egg)
    e_differences[i,j] = diff
  }
}
```

Calculate phenotype drift

```{r}
# same for phenotype 
n = length(df_f[,1])
f_differences = matrix(ncol=n, nrow=n)

for ( i in c(1:n)){
  current = as.numeric(df_f[i,])
  for ( j in c(i:n)){
    egg = as.numeric(df_f[j,])
    diff = mse(current, egg)
    f_differences[i,j] = diff
  }
}

```


Plot drfit heatmap

```{r}
pal <- colorRampPalette(c("red", "yellow"), space = "rgb")
levelplot(t(e_differences), main="DNA MSE", xlab="", ylab="", col.regions=pal(100))
# levelplot(e_differences, scales=list(x=list(rot=-90)),main="DNA MSE", xlab="", ylab="", col.regions=pal(100))

pal <- colorRampPalette(c("red", "yellow"), space = "rgb")
levelplot(t(f_differences), main="Phenotype MSE", xlab="", ylab="", col.regions=pal(100))

```

Saving / Loading files

```{r}
# save to file
fileName = paste(folder_name, "fish_dna_diff_matrix_", max_step, ".csv", sep="")
write.table(e_differences, fileName, quote=FALSE)

fileName = paste(folder_name, "phenotype_diff_matrix_", max_step, ".csv", sep="")
write.table(f_differences, fileName, quote=FALSE)
```

```{r}
# load files

max_step= 19200
fileName = paste(folder_name, "fish_dna_diff_matrix_", max_step, ".csv", sep="")
e_differences = data.matrix(read.table(fileName, sep = " "))

fileName = paste(folder_name, "phenotype_diff_matrix_", max_step, ".csv", sep="")
f_differences = data.matrix(read.table(fileName, sep = " "))
```


Plot the difference of each generation relative to generation 0

```{r}
a = e_differences[1,]
plot(a, xlab = "Generation", ylab = "DNA MSE", cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
b = f_differences[1,]
plot(b, xlab = "Generation", ylab = "Phenotype MSE",
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
plot(x=a, y = b[1:length(b)-1], xlab = "DNA MSE", ylab = "Phenotype MSE",
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)

```

Plot genetic drift and phenotype drift

```{r}
mean_ph_diff = average_difference(f_differences)
mean_dna_diff = average_difference(e_differences)

plot(mean_ph_diff, xlab = "Xth descendant", ylab = "Average phenotype MSE",
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5) 
plot(mean_dna_diff, xlab = "Xth descendant", ylab = "Average DNA MSE",
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5) 
# correlation, genotype-phenotype mapping
plot(x=mean_dna_diff, y = mean_ph_diff[1:length(mean_ph_diff)-1],
      xlab = "Average DNA MSE", ylab = "Average phenotype MSE",
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)

```

```{r}
# correlation, genotype-phenotype mapping
plot(x=t(mean_dna_diff[2:n]), y = t(mean_ph_diff[2:n]))
```

```{r}
stats_func <- function(x) c(sum=sum(x), sdev=var(x)) 
```

Plot the drift with error bars (not exactly helpful so not included in paper)

```{r}
#  shift values
df = f_differences
n = length(df[,1])
# sum after moving the row back by one column
for (i in c(2:n)){
  sub_ = df[i,i:n]
  df[i, ] = c(sub_, rep(0, n-length(sub_)))
}

avg = apply(df, 2, sum)/c(n:1)

# do it again for standard error
var = rep(0,n)
for (i in c(1:n)){
  sub_ = df[1:(n-i),i] - avg[i]
  var[i] = sum(sub_*sub_)/(n-i+1)
  var[i] = sqrt(var[i]/(n-i+1))
}


plot(avg)
plot(var)
sdev = var

x <- 1:n
plot(x, avg,
    ylim=range(c(avg-sdev, avg+sdev)),
    xlab = "Xth descendant", ylab = "Average phenotype MSE",
    cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5
)
# hack: we draw arrows but with very special "arrowheads"
arrows(x, avg-sdev, x, avg+sdev, length=0.05, angle=90, code=3)

```

----- Miscellanous functions

Plot organism image from data array

```{r}
df_to_img = function(df_vect, h, w){
  # rgb, a = x[..., :3], clip_by_value(x[..., 3:4], 0.0, 1.0)
  # return 1.0-a+rgb
  
  
  egg = as.numeric(df_vect[1,])
  egg = matrix(egg, 16, h*w)
  egg = t(egg)
  rgb = egg[1:(h*w), 1:3]
  a = pmax( 0, pmin( egg[1:(w*h), 4], 1))
  egg_rgb = rgb + (1.0-a)
  # at this point it's -1:+1
  # egg_rgb = (1+egg_rgb)/2
  egg_rgb = pmax( 0, pmin(egg_rgb, 1))
  
  data <- array(data = egg_rgb, dim = c(w,h,3)) #c(w,h,3)
  
  # t(data) is not directly possible
  tdata =  array(data = egg_rgb, dim = c(h,w,3)) #c(w,h,3)
  for (i in c(1:w)){
    tdata[,i,] = data[i,,]
  }
  
  plotArrayAsImage(tdata, main = "")
}
```


Save image

```{r}
save = TRUE

output_path = paste(folder_name, "additional_gen/", sep="")
dir.create(file.path(output_path), showWarnings = FALSE)

for (i in c(80:100)){
  filename = paste(i, ".png", sep="")
  if(save){
    png(paste(output_path, filename, sep=""), units="mm", width=200, height=150, res=300)
  }
  df_to_img(df_f[i,], 40 ,40)
  if(save){
    dev.off()
  }
}


```

